---
title: "Estimating Discrete Choice Models using Apollo"
author: "Bryan Parthum"
date: "`r format(Sys.Date(), format='%B %e, %Y')`"
output:
  bookdown::html_document2:
    df_print: paged
    toc: true
    toc_float: true
    fig_caption: yes
    depth: 2
    number_sections: true
    code_folding:  hide
    theme: readable
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(magrittr)
library(kableExtra)
library(apollo)
```

This document walks a user through estimation of a discrete choice model using the [Apollo](http://www.apollochoicemodelling.com/index.html) package in [Rstudio](https://www.rstudio.com/). The example code is drawn directly from the helpful list of [examples](http://www.apollochoicemodelling.com/examples.html) found on the Apollo website.

# Simple Multinomial Logit (MNL) model on mode choice SP data

Start by initializing the code and setting up core controls.

```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}

## initialize code
apollo_initialise()

## set core controls
apollo_control = list(
  modelName       = "MNL_SP",
  modelDescr      = "Simple MNL model on mode choice SP data",
  indivID         = "ID",
  outputDirectory = "output"
)
```

Next, load the example data and subset it to only include the stated preference data (SP==1). For a dictionary of available data, use `? apollo_modeChoiceData`. If necessary, transform the data to be readable by Apollo. 

```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}
## load data
database = apollo_modeChoiceData %>% 
  filter(SP==1)
```

Determine the attributes of the model and define the priors on the model parameters. If none are available, a prior of `0` should be used. Define any parameters (using the names of the parameters) that should be fixed at their starting value.

```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}
## vector of parameters, including any that are kept fixed in estimation
apollo_beta = c(
  asc_car      = 0,
  asc_bus      = 0,
  asc_air      = 0,
  asc_rail     = 0,
  b_tt_car     = 0,
  b_tt_bus     = 0,
  b_tt_air     = 0,
  b_tt_rail    = 0,
  b_access     = 0,
  b_cost       = 0,
  b_no_frills  = 0,
  b_wifi       = 0,
  b_food       = 0
)

## vector with names (in quotes) of parameters to be kept fixed at their starting value in params, use apollo_beta_fixed = c() if none
apollo_fixed = c("asc_car","b_no_frills")
```

Group and validate the inputs of the model using the functions defined within Apollo. 

```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}
## group and validate
apollo_inputs = apollo_validateInputs(apollo_beta,
                                      apollo_fixed,
                                      database)
```

Define the model and the likelihood function.

```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}
## likelihood function
apollo_probabilities=function(apollo_beta, 
                              apollo_inputs, 
                              functionality="estimate"){
  
  ## attach inputs and detach after function exit
  apollo_attach(apollo_beta, apollo_inputs)
  on.exit(apollo_detach(apollo_beta, apollo_inputs))
  
  ## create list of probabilities P
  P = list()
  
  ## list of utilities: these must use the same names as in mnl_settings, order is irrelevant
  V = list()
  V[["car"]]  = asc_car + b_tt_car * time_car + b_cost * cost_car
  V[["bus"]]  = asc_bus + b_tt_bus * time_bus + b_access * access_bus + b_cost * cost_bus
  V[["air"]]  = asc_air + b_tt_air * time_air + b_access * access_air + b_cost * cost_air + b_no_frills * ( service_air == 1 ) + b_wifi * ( service_air == 2 ) + b_food * ( service_air == 3 )
  V[["rail"]] = asc_rail + b_tt_rail * time_rail + b_access * access_rail + b_cost * cost_rail + b_no_frills * ( service_rail == 1 ) + b_wifi * ( service_rail == 2 ) + b_food * ( service_rail == 3 )
  
  ## define settings for MNL model component
  mnl_settings = list(
    alternatives  = c(car = 1, bus = 2, air = 3, rail = 4), 
    avail         = list(car = av_car, bus = av_bus, air = av_air, rail = av_rail), 
    choiceVar     = choice,
    utilities     = V
  )
  
  ## compute probabilities using MNL model
  P[["model"]] = apollo_mnl(mnl_settings, functionality)
  
  ## take product across observation for same individual
  P = apollo_panelProd(P, apollo_inputs, functionality)
  
  ## prepare and return outputs of function
  P = apollo_prepareProb(P, apollo_inputs, functionality)
  return(P)
}
```

We are now ready to estimate the model.

```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}
## estimate
mnl = apollo_estimate(apollo_beta, apollo_fixed, apollo_probabilities, apollo_inputs)
```

Explore model results. 
```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}
## model outputs
apollo_modelOutput(mnl)

## optional: save model output to output directory
# apollo_saveOutput(model)
```

# Mixed logit model (MXL) on Swiss route choice data, correlated Lognormals in utility space

This model is significantly more computationally intensive than the previous simple model. To speed up the estimation, first check to see how many cores are available on your machine. The number of cores chosen in this example is `14` using the option `nCores = 14`. Initialize the code and set up core controls. 
```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}

## initialize code
apollo_initialise()

## set core controls
apollo_control = list(
  modelName       = "MXL_preference_space_correlated",
  modelDescr      = "Mixed logit model on Swiss route choice data, correlated Lognormals in utility space",
  indivID         = "ID",  
  mixing          = TRUE, 
  nCores          =  14,
  outputDirectory = "output"
)
```

Next, load the example data. For a dictionary of available data, use `? apollo_modeChoiceData`. If necessary, transform the data to be readable by Apollo. 

```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}
## load data
database = apollo_swissRouteChoiceData
```

Determine the attributes of the model and define the priors on the model parameters (and their standard deviations). If none are available, a prior of `0` should be used. Define any parameters (using the names of the parameters) that should be fixed at their starting value.

```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}
## vector of parameters, including any that are kept fixed in estimation
apollo_beta = c(
  mu_log_b_tt       = -3,
  sigma_log_b_tt    =  0,
  mu_log_b_tc       = -3,
  sigma_log_b_tt_tc =  0,
  sigma_log_b_tc    =  0,
  mu_log_b_hw       = -3,
  sigma_log_b_tt_hw =  0,
  sigma_log_b_tc_hw =  0,
  sigma_log_b_hw    =  0,
  mu_log_b_ch       = -3,
  sigma_log_b_tt_ch =  0,
  sigma_log_b_tc_ch =  0,
  sigma_log_b_hw_ch =  0,
  sigma_log_b_ch    =  0
)

## vector with names (in quotes) of parameters to be kept fixed at their starting value in params, use apollo_beta_fixed = c() if none
apollo_fixed = c()

## if they exist, read in starting values for at least some parameters from existing model output file
# apollo_beta=apollo_readBeta(apollo_beta, apollo_fixed, "MXL_preference_space_correlated", overwriteFixed = FALSE)
```

Define random components.

```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}

apollo_draws = list(
  interDrawsType = "halton",
  interNDraws    = 500,
  interUnifDraws = c(),
  interNormDraws = c("draws_tt","draws_tc","draws_hw","draws_ch"),
  intraDrawsType = "halton",
  intraNDraws    = 0,
  intraUnifDraws = c(),
  intraNormDraws = c()
)
```

Create random parameters.

```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}

## random parameters
apollo_randCoeff = function(apollo_beta, apollo_inputs){
  randcoeff = list()

  randcoeff[["b_tt"]] = -exp( mu_log_b_tt  + sigma_log_b_tt * draws_tt )
  randcoeff[["b_tc"]] = -exp( mu_log_b_tc  + sigma_log_b_tt_tc * draws_tt + sigma_log_b_tc * draws_tc )
  randcoeff[["b_hw"]] = -exp( mu_log_b_hw  + sigma_log_b_tt_hw * draws_tt + sigma_log_b_tc_hw * draws_tc + sigma_log_b_hw * draws_hw )
  randcoeff[["b_ch"]] = -exp( mu_log_b_ch  + sigma_log_b_tt_ch * draws_tt + sigma_log_b_tc_ch * draws_tc + sigma_log_b_hw_ch * draws_hw + sigma_log_b_ch * draws_ch )

  return(randcoeff)
}
```

Group and validate the inputs of the model.

```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}
## group and validate
apollo_inputs = apollo_validateInputs()
```

Define the model and the likelihood function.

```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}
## likelihood function
apollo_probabilities=function(apollo_beta, apollo_inputs, functionality="estimate"){
  
  ## function initialization: do not change the following three commands
  ## attach inputs and detach after function exit
  apollo_attach(apollo_beta, apollo_inputs)
  on.exit(apollo_detach(apollo_beta, apollo_inputs))

  ## create list of probabilities P
  P = list()
  
  ## list of utilities: these must use the same names as in mnl_settings, order is irrelevant
  V = list()
  V[["alt1"]] = b_tt * tt1 + b_tc * tc1 + b_hw * hw1 + b_ch * ch1
  V[["alt2"]] = b_tt * tt2 + b_tc * tc2 + b_hw * hw2 + b_ch * ch2
  
  ## define settings for MNL model component
  mnl_settings = list(
    alternatives  = c(alt1=1, alt2=2),
    avail         = list(alt1=1, alt2=1),
    choiceVar     = choice,
    utilities     = V
  )
  
  ## compute probabilities using MNL model
  P[["model"]] = apollo_mnl(mnl_settings, functionality)
  
  ## take product across observation for same individual
  P = apollo_panelProd(P, apollo_inputs, functionality)
  
  ## average across inter-individual draws
  P = apollo_avgInterDraws(P, apollo_inputs, functionality)
  
  ## prepare and return outputs of function
  P = apollo_prepareProb(P, apollo_inputs, functionality)
  return(P)
}

```

We are now ready to estimate the model.

```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}
## estimate
mxl = apollo_estimate(apollo_beta, apollo_fixed,
                        apollo_probabilities, apollo_inputs)
```

Explore model results. 
```{r  include=TRUE, message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, class.source="fold-show"}
## model outputs
apollo_modelOutput(mxl)

## optional: save model output to output directory
# apollo_saveOutput(model)
```
